Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2009-08-28 docs/pubs/ISEI6/response.txt

----------------------------------------------------------------------
5). "In the references: Usually, the word "pages" is abbreviated to
pp. Ref 18: should be to a journal article, not a web site - perhaps
Raskin and Pan (2005) in Computers & Geosciences. Ref 19: to EKAW?
What is that?  Compare Refs 4 and 7 - different wordings are used to
describe the same conference."

FIXED

----------------------------------------------------------------------
Reviewer #2:
----------------------------------------------------------------------
6). "In the early stages of the paper there are some assertions
without reference about what is and is not difficult. See p.2 end of
2nd, 3rd and 4th paragraphs.  Can a little more detail be added to
support the statements? If not, then noting that these are
suppositions would be useful for the reader."

To address this issue we added additional text on the need for new
tools in ontology development for ecologists (e.g., compared with ...)
Also, much of the motivation for the work is based on our work with
ecologists and information managers, which we refer to in the paper.

@@

----------------------------------------------------------------------
7). "Was there any quantifiable basis for choosing the predicates?  In
particular, the statement" "While not as expressive as OWL-DL, our
approach can produce ontology structures that are essential for
improved data discovery and integration [14]."  indicates some level
of decision in the reduced description but without any details,
especially in terms for DL-expressivity, it is SHIQ, SHOIN, or what?"

As far as I am aware, the arrangement of DL constructs used in
OWLifier does not directly map to any previously named DL sublanguage,
however, it does provides only OWL-DL based constructs, and thus can
be considered best as a sublanguage of SHOIN(D). Note that although
many of the OWL-DL constructs can be used, they can only be used in
certain contexts. Also, some of the more foundational constructs of
DLs are not considered, e.g., value restrictions and concept
disjunction. While additional blocks can easily be added to OWLifier
spreadsheet conventions, the proposed set of blocks were chosen to
provide a simplified syntax (for a novice DL user) while leading to a
"natural" or "intuitive" semantics of blocks for these users.

----------------------------------------------------------------------
8). "There are no mention of instances, that I could find, at all of
instances?  Are these part of the use case or not? If not, this really
needs to be stated.  An ontology without instances has limited
application."

Our main goal is to provide a mechanism for describing ecological
concepts, and then to allow additional tools to use these concepts for
annotating scientific data (which represents the instances/values). We
briefly mention this now in the introduction, and also elaborate on
the scope of owlifier later in the paper.

@@

[josh]: Maybe at the start of third paragraph of introduction: This
paper presents a novel approach for ontology creation that aims at
being more intuitive for ecologists that can be used to rapidly
construct large ontologies of ecological terms (classes) and the
subsequent annotation of scientific data (instances).

----------------------------------------------------------------------
9). "Fig.1 suggests a round-trip approach to knowledge engineering,
but in fact it is not. How does the present formulation represent to
scientiest what gets dropped in the back conversion from OWL to the
spreadsheet format?"

We reworded the implementation section to make it clearer that the
conversion is not information preserving. ... the owlifier tool will
notify the user of this situation, but at this time no additional
support is provided.

### 2009-08-28 docs/pubs/ISEI6/main.tex
increment

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop broad and consistent terminologies within ecology
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. For instance, annotating data with
ontology terms can both help users interpret data as well as enable
advanced capabilities for data discovery and integration, e.g., by
exploiting subsumption and part-of hierarchies as well as more formal
constraints such as cardinality restrictions on properties and term
equivalence \cite{madin08:_advan_ecolog_resear_with_ontol}.

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax. We see the lack of suitable ontology editing tools for
scientists without expertise in knowledge representation as one of the
major barriers for more wide-scale adoption of ontologies in ecology.

This paper presents a novel approach for ontology creation that aims
at being more intuitive for ecologists and that can be used to rapidly
construct large ontologies for describing scientific data.  Our
approach is to allow scientists to use common and familiar
spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
Office Spreadsheet) to describe, in an intuitive way, different
aspects of an ontology, and then to take these descriptions and
convert them into full-fledged OWL ontologies using a software
application called \owlifier. The use of a familiar editing
environment can improve the speed and understanding of ontology
construction and avoid locating, downloading, and interpreting
additional software packages and add-ons (e.g., CMAP-COE
\cite{hayes05:_collab_knowl_captur_in_ontol}).  Many spreadsheet
applications also contain easy-to-use features that support scientists
with the structuring process (e.g., cut/copy/paste, search/replace,
sort, split-window, annotation, track-changes, freeze panes, and so
on).

@@

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct relatively complex and
meaningful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise)
\cite{rector04:_owl_pizzas}. In \secref{sec:features} we describe
additional characteristics of \owlifier\ and discuss issues with
respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the \owlifier\
implementation, and conclude in \secref{sec:conclusion} with related
and future work. In general, the goal of \owlifier\ is not to support
all constructs in OWL-DL, but instead to provide a higher-level
ontology syntax (via spreadsheet blocks) that is easy for ecologists
to use and understand while also providing the necessary constructs
for developing typical ecological ontologies. By compiling \owlifier\
to OWL-DL, we also allow for experts to refine and extend the ontology
using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

As described above, an \owlifier\ table defines an OWL-DL
\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. 
% This section describes in detail each type of block supported by
% \owlifier.
We assume that if any properties or classes used in a block are not
imported from another ontology, then they are to be added to the
ontology being specified by the \owlifier\ table (i.e., the
``current'' ontology). In general, we name blocks according to the
terms used in
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
as opposed to the names used for corresponding constructs in
OWL-DL. This choice of block names helps to simplify terminology
(e.g., we use ``relation'' below instead of ``object property''),
allows \owlifier\ to easily generate ontologies that extend the
observational model of
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
and avoids confusion with established terms commonly used within
ecology (e.g., ``class'').

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \textsf{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. As an example, the following block imports the SWEET
``Earth Realm'' ontology \cite{raskin05}
\begin{small}
\begin{tabbing}
  ~~\textsf{import}  \textsf{sweet} 
  \textsf{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
\end{small}
With this import block the class denoting Marine Ecosystems (a class
defined in the SWEET ontology) can be referred to from within an
\owlifier\ table using the expression
\textsf{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

@@

% \myblock{Value Blocks.} Value blocks define required datatype property
% \emph{constant values} for concepts. A value block has the form
% \begin{itemize}
% \item[]
%   \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{value} & $P$ & $V$ & $C_1$
%     & $C_2$ & \dots & $C_n$ \\ \hline
%   \end{tabular} \hfill ($n \ge 1$)
% \end{itemize}
% where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
% n$, and $V$ is a datatype value. For each concept $C_i$, the value
% block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
% each instance of $C_i$ has a value $V$ for property $P$.  The value
% restrictions stated by value blocks are often used for defining
% so-called \emph{value partitions} \cite{co-ode}.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes.  In particular, unless
explicitly asserted, distinct classes within an OWL-DL ontology are
not assumed to be disjoint. However, in many ontologies a large number
of classes are typically defined as being disjoint (e.g., sibling
classes), and stating these disjoint constraints is often
time consuming since each pair of classes must be given an explicit
disjoint assertion. Editors such as Protege
\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
via the user interface to create specific sets of disjoint assertions,
e.g., by allowing a user to define all children of a particular class
as disjoint. In general, however, many users expect such classes to be
disjoint by default \cite{rector04:_owl_pizzas} and this expectation
often leads to modeling errors.

Alternatively, the default assumption in \owlifier\ is that distinct
classes are disjoint. Specifically, as an \owlifier\ table is
converted to an OWL-DL ontology, the system analyzes the class
hierarchy structure and identifies pairs of classes that are: (1) not
related via subclass relations (either direct or indirect subclasses);
(2) not defined as synonyms; and (3) not explicitly defined to overlap
via an overlap block. Each such pair of classes is then asserted by
\owlifier\ in the resulting ontology as being disjoint. As described
in \cite{rector04:_owl_pizzas}, undeclared disjoint classes are a
common problem in ontology development using OWL-DL and often limit
the utility of reasoning systems (by limiting the inferences that can
be obtained). The approach employed in \owlifier\ for handling
disjoint classes makes the common expectations of users the default
case, which in general should lead to a more intuitive ontology
editing environment and an overall fewer number of modeling mistakes.

@@

As described in \cite{rector04:_owl_pizzas} additional reasoning can
occur within OWL-DL ontologies when domain and range axioms are
provided (as well as, e.g., property closure axioms).  We explicitly
do not consider these constraints in the current version of \owlifier\
because they also often result in modeling errors for inexperienced
OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the
approach of more traditional description logics, which do not have
explicit domain and range axioms.  Although not currently supported in
\owlifier, domain and range constraints as well as property closure
axioms can be inferred from given relationship blocks.

@@

To help simplify the creation of class hierarchies and property
sequences (including transitive, cardinality, and sufficient blocks),
we allow for ``sparse'' blocks that inherit missing information from
their closest proceeding block. For instance the following entity
blocks
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
  \textsf{Fjord} \\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{TidalMarsh} \\ 
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{SaltMarsh}
\end{tabbing}
can be equivalently represented in \owlifier\ using the following
sparse blocks.
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  \> \> \> \textsf{Fjord} \\
  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
  \> \> \> \textsf{SaltMarsh}
\end{tabbing}
In general, the use of sparse blocks simplifies ontology creation by
not requiring users to enter every redundant field explicitly, which
in turn can simplify the overall layout of the ontology within a
spreadsheet.  Additionally, \owlifier\ does not place constraints on
the order of blocks within a spreadsheet. Classes also do not need to
be explicitly defined within an entity block, e.g., classes without
corresponding entity blocks can be introduced simply through synonym
and relationship blocks. This typically occurs when a particular class
does not participate as a subclass or superclass of another class in
the current ontology. Taken together, these appraoches allow users to
enter minimal information by limiting redundancy and by supporting a
range of inferences, while at the same time reducing the causes of
many errors commonly made in ontology development by non-experts.

@@

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \cite{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\cite{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While straightforward to perform, this last
conversion is not information preserving since not all OWL-DL
constructs are represented through \owlifier\ blocks.

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the (potentially lossy) back conversion from OWL-DL files to
corresponding \owlifier\ blocks. The current implementation of
\owlifier\ is being used within a project focused on integrating
vegetation trait data.  As noted above, not all the blocks described
here are yet supported, however, these priority needs are clearly
emerging from our interactions with the vegetation scientists.
Nevertheless, these scientists were able to rapidly prototype and
revise their trait ontologies using \owlifier, with little to no
instruction in formal logic or knowledge modeling.  In addition,
\owlifier\ is being used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations between existing ontologies and the observation
ontology defined in \cite{madin07:_ontol_for_descr_and_synth}.

@@

Protege provides a variety of ontology editing plug-ins, including a
simple interface for text-based editing of class hierarchies. In
\cite{kola07:_impor_spread_data_into_proteg}, an approach is described
for importing spreadsheet-based ontology descriptions into
Protege. However, this approach aims at supporting \emph{ad-hoc}
spreadsheet structures by providing an intermediate interface for
mapping these structures into ontology axioms. This approach is
similar to others (e.g.,
\cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
for defining mappings from relational data to RDF and OWL ontology
class and instance data. To the best of our knowledge, however, our
approach is the first to consider a detailed set of templates that can
support a large subset of existing OWL constructs. In addition, we
define a number of shortcuts for creating \owlifier\ tables, including
sparse blocks and default semantics (e.g., disjoint classes, which is
a form of global taxonomic constraint defined in
\cite{thau07:_reason_about_taxon_in_first_order_logi}) that further
simplify ontology creation for end users.

>>>>

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop broad and consistent terminologies within ecology
\citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. For instance, annotating data with
ontology terms can both help users interpret data as well as enable
advanced capabilities for data discovery and integration, e.g., by
exploiting subsumption and part-of hierarchies as well as more formal
constraints such as cardinality restrictions on properties and term
equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\citep{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax.  A well-funded ontology development effort, the Gene Ontology
(GO), has been underway for over a decade within the molecular biology
community, and has led to some major successes enriching genomic and
biomedical data resources with annotations to community-based
ontologies. A major lesson learned from these efforts was the
importance of keeping the ontology focused, and indeed GO has a highly
delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}
Ecology, on the other hand, encompasses an extremely broad range of
scales and disciplines, such that a coordinated ontology development
effort would need to involve a large number of experts to cover the
diverse types of information relevant to ecological investigations.
At this point, we believe that ecological researchers need to begin
constructing ontologies for their own specialized subdisciplines.
Most ontology editing tools, however, typically require substantial
expertise in formal knowledge representation, and present a major
barrier for more wide-scale creation and adoption of ontologies in
ecology.

This paper presents a novel approach for ontology creation that aims
at being more intuitive for ecologists and that can be used to rapidly
construct large ontologies of ecological terms (classes) and the
subsequent annotation of scientific data (instances).  Our approach is
to allow scientists to use common and familiar spreadsheet-based tools
(e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to
describe, in an intuitive way, different aspects of an ontology, and
then to take these descriptions and convert them into full-fledged OWL
ontologies using a software application called \owlifier. The use of a
\emph{familiar} editing environment can improve the speed and
understanding of ontology construction and can help users avoid the
often time-consuming task of locating, downloading, installing, and
learning fundamentally new software applications and interfaces (e.g.,
CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
\citep{knublauch04:_editin_descr_logic_ontol_with}).  Many spreadsheet
applications also contain easy-to-use features that support scientists
in the process of editing and managing \owlifier\ ontology documents
(e.g., via cut/copy/paste, search/replace, sort, split-window,
track-changes, freeze panes, and so on).

@@

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct relatively complex and
meaningful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\citep{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise)
\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
additional characteristics of \owlifier\ and discuss issues with
respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the \owlifier\
implementation, and conclude in \secref{sec:conclusion} with related
and future work. In general, the goal of \owlifier\ is not to support
all constructs in OWL-DL, but instead to provide a higher-level
ontology syntax (via spreadsheet blocks) that is easy for ecologists
to use and understand while also providing the necessary constructs
for developing typical ecological ontologies. By compiling \owlifier\
to OWL-DL, we also allow for experts to refine and extend the ontology
using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

As described above, an \owlifier\ table defines an OWL-DL
\citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. 
% This section describes in detail each type of block supported by
% \owlifier.
We assume that if any properties or classes used in a block are not
imported from another ontology, then they are to be added to the
ontology being specified by the \owlifier\ table (i.e., the
``current'' ontology). In general, we name blocks according to the
terms used in
\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
as opposed to the names used for corresponding constructs in
OWL-DL. This choice of block names helps to simplify terminology
(e.g., we use ``relation'' below instead of ``object property''),
allows \owlifier\ to easily generate ontologies that extend the
observational model of
\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
and avoids confusion with established terms commonly used within
ecology (e.g., ``class'').

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \textsf{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. As an example, the following block imports the SWEET
``Earth Realm'' ontology \citep{raskin05}
\begin{small}
\begin{tabbing}
  ~~\textsf{import}  \textsf{sweet} 
  \textsf{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
\end{small}
With this import block the class denoting Marine Ecosystems (a class
defined in the SWEET ontology) can be referred to from within an
\owlifier\ table using the expression
\textsf{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

@@

% \myblock{Value Blocks.} Value blocks define required datatype property
% \emph{constant values} for concepts. A value block has the form
% \begin{itemize}
% \item[]
%   \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{value} & $P$ & $V$ & $C_1$
%     & $C_2$ & \dots & $C_n$ \\ \hline
%   \end{tabular} \hfill ($n \ge 1$)
% \end{itemize}
% where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
% n$, and $V$ is a datatype value. For each concept $C_i$, the value
% block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
% each instance of $C_i$ has a value $V$ for property $P$.  The value
% restrictions stated by value blocks are often used for defining
% so-called \emph{value partitions} \citep{co-ode}.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\citep{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes.  In particular, unless
explicitly asserted, distinct classes within an OWL-DL ontology are
not assumed to be disjoint. However, in many ontologies a large number
of classes are typically defined as being disjoint (e.g., sibling
classes), and stating these disjoint constraints is often
time consuming since each pair of classes must be given an explicit
disjoint assertion. Editors such as Protege
\citep{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
via the user interface to create specific sets of disjoint assertions,
e.g., by allowing a user to define all children of a particular class
as disjoint. In general, however, many users expect such classes to be
disjoint by default \citep{rector04:_owl_pizzas} and this expectation
often leads to modeling errors.

Alternatively, the default assumption in \owlifier\ is that distinct
classes are disjoint. Specifically, as an \owlifier\ table is
converted to an OWL-DL ontology, the system analyzes the class
hierarchy structure and identifies pairs of classes that are: (1) not
related via subclass relations (either direct or indirect subclasses);
(2) not defined as synonyms; and (3) not explicitly defined to overlap
via an overlap block. Each such pair of classes is then asserted by
\owlifier\ in the resulting ontology as being disjoint. As described
in \citep{rector04:_owl_pizzas}, undeclared disjoint classes are a
common problem in ontology development using OWL-DL and often limit
the utility of reasoning systems (by limiting the inferences that can
be obtained). The approach employed in \owlifier\ for handling
disjoint classes makes the common expectations of users the default
case, which in general should lead to a more intuitive ontology
editing environment and an overall fewer number of modeling mistakes.

@@

As described in \citep{rector04:_owl_pizzas} additional reasoning can
occur within OWL-DL ontologies when domain and range axioms are
provided (as well as, e.g., property closure axioms).  We explicitly
do not consider these constraints in the current version of \owlifier\
because they also often result in modeling errors for inexperienced
OWL-DL users \citep{rector04:_owl_pizzas}. Instead, we adopt the
approach of more traditional description logics, which do not have
explicit domain and range axioms.  Although not currently supported in
\owlifier, domain and range constraints as well as property closure
axioms can be inferred from given relationship blocks.

@@

To help simplify the creation of class hierarchies and property
sequences (including transitive, cardinality, and sufficient blocks),
we allow for ``sparse'' blocks that inherit missing information from
their closest proceeding block. For instance the following entity
blocks
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
  \textsf{Fjord} \\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{TidalMarsh} \\ 
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{SaltMarsh}
\end{tabbing}
can be equivalently represented in \owlifier\ using the following
sparse blocks.
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  \> \> \> \textsf{Fjord} \\
  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
  \> \> \> \textsf{SaltMarsh}
\end{tabbing}
In general, the use of sparse blocks simplifies ontology creation by
not requiring users to enter every redundant field explicitly, which
in turn can simplify the overall layout of the ontology within a
spreadsheet.  Additionally, \owlifier\ does not place constraints on
the order of blocks within a spreadsheet. Classes also do not need to
be explicitly defined within an entity block, e.g., classes without
corresponding entity blocks can be introduced simply through synonym
and relationship blocks. This typically occurs when a particular class
does not participate as a subclass or superclass of another class in
the current ontology. Taken together, these approaches allow users to
enter minimal information by limiting redundancy and by supporting a
range of inferences, while at the same time reducing the causes of
many errors commonly made in ontology development by non-experts.

@@

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \citep{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While straightforward to perform, this back
conversion is not information preserving since not all OWL-DL
statements can be represented through \owlifier\ blocks.

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the (potentially lossy) back conversion from OWL-DL files to
corresponding \owlifier\ blocks. The current implementation of
\owlifier\ is being used within a project focused on integrating
vegetation trait data.  As noted above, not all the blocks described
here are yet supported, however, these priority needs are clearly
emerging from our interactions with the vegetation scientists.
Nevertheless, these scientists were able to rapidly prototype and
revise their trait ontologies using \owlifier, with little to no
instruction in formal logic or knowledge modeling.  In addition,
\owlifier\ is being used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations between existing ontologies and the observation
ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.

@@

Protege provides a variety of ontology editing plug-ins, including a
simple interface for text-based editing of class hierarchies. In
\citep{kola07:_impor_spread_data_into_proteg}, an approach is described
for importing spreadsheet-based ontology descriptions into
Protege. However, this approach aims at supporting \emph{ad-hoc}
spreadsheet structures by providing an intermediate interface for
mapping these structures into ontology axioms. This approach is
similar to others (e.g.,
\citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
for defining mappings from relational data to RDF and OWL ontology
class and instance data. To the best of our knowledge, however, our
approach is the first to consider a detailed set of templates that can
support a large subset of existing OWL constructs. In addition, we
define a number of shortcuts for creating \owlifier\ tables, including
sparse blocks and default semantics (e.g., disjoint classes, which is
a form of global taxonomic constraint defined in
\citep{thau07:_reason_about_taxon_in_first_order_logi}) that further
simplify ontology creation for end users.

As future work, we intend to extend our current \owlifier\
implementation to support the full set of blocks defined above. We
also intend to explore approaches to fully support conversions between
\owlifier\ tables and OWL-DL ontologies. In particular, we want to
extend \owlifier\ to re-apply changes made by knowledge representation
experts that previously modified the OWL-DL ontology generated by
\owlifier. For instance, if an OWL-DL ontology generated from an
\owlifier\ table is refined and extended by a knowledge representation
expert, then converted back into an \owlifier\ ontology that is
further edited and extended by a scientist, and then converted again
into an OWL-DL ontology, we want to maintain (i.e., re-apply) the
original extensions and edits created by the knowledge representation
expert that are still relevant. We are also developing extensions to
\owlifier\ to support conversion into the observation ontology
framework presented in \citep{bowers08:_concep_model_framew_for_expres}
and intend to perform additional testing and evaluation of the
\owlifier\ approach with ecologists.
